Local Type Inference for Polarised System F with Existentials

### 2024-02-13 oopsla24/tex/02-overview.tex
wip

\subsection{Translation to \fexists}

@@

To establish a correspondence between \fexists and standard unpolarized systems,
we present a translation between \fexists and \systemf. The translation of
\systemf to \fexists is given in \cref{fig:translation}.

\begin{figure}
  \begin{equation*}
  \begin{aligned}
    [[|↓iN|]] &= [[|iN|]] \\
    [[|↑iP|]] &= [[|iP|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|α±|]] &= [[a]] \\
    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type depolarization}
  \label{fig:depolarization}
\end{figure}

\begin{figure}
  \begin{equation*}
  \begin{aligned}
    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type polarization}
  \label{fig:polarization}
\end{figure}

\subsection{What types do we infer?}

Suppose that
we have the following variables in context $[[Φ]]$:

\begin{align*}
  [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
  [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
  [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
  [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
  [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
\end{align*}

Then

\begin{align*}
  &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
  &[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]] \ilyam{(no)}\\
\end{align*}

@@

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similarly to the \CBPV system \cite{levy2006:cbpv}.
The negative types represent computations, and the positive types represent values:
\begin{itemize}
\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables 
  $[[pas]]$. The polarities are chosen to follow the definition of functions.
\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$.
\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
  binds negative variables in a positive type $[[iP]]$. 
\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
\end{itemize}

@@

>>>>

% Imprediicative
% CBPV, monadic control
% Local Inference
% Two ways to translate

@@

Positive expressions correspond to values in the \CBPV system
\cite{levy2006:cbpv}, and negative expressions correspond to computations.
The difference between the computations and values is intuitively described 
in the slogan of \CBPV: ``a value is, a computation does''. 
In particular, an argument of a function is always a value (\ie positive), 
and the resulting type, as well as the function itself, 
is a computation (\ie negative).

\subsection{Examples}

The standard \systemf types and terms must be polarized to be used in \fexists, 
we present the embedding formally in \cref{fig:polarization}.
For instance, the following types and terms of \systemf 
can be polarized in the given way:

\begin{figure}[h]
  \begin{align*}
    [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
    [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
    [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
    [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
    [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
    [[autoP]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
  \end{align*}
  \caption{Polarization of standard \systemf terms}
  \label{fig:polarization-examples}
\end{figure}

Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
Then the following types are inferrable in \fexists:

\begin{align*}
  % &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
  &[[· ; Φ, x1:↓(↓iN → iN), x2:↓(↓iM → iM) ⊢ let x = choose (x1, x2); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]\\
  &[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]\\
  &[[· ; Φ ⊢ let x = choose (id, autoP); return x : ↑ ∃γ⁻.↓γ⁻]]\\
\end{align*}

The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
which has certain limitations.
In particular, if an argument of a function is polymorphic,
its polymorphic variables are not instantiated by the inference algorithm. 
For example, one would expect the following to be inferrable:
$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$.
However, to infer this, the $[[id]]$ must be instantiated to $[[Int → ↑Int]]$, 
which requires the information to be propagated from the neighboring branch 
of the syntax tree ($[[ [two, three, nine] ]]$), \ie bypass the locality.

However, the polymorphic arguments of the function are instantiated, 
and thus, the annotated version of the same term infers the type successfully:
$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$.

@@

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similar to the values and computations in the \CBPV system \cite{levy2006:cbpv}.
\begin{itemize}
\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables 
  $[[pas]]$. The polarities are chosen to follow the definition of functions.
\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$.
\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
  binds negative variables in a positive type $[[iP]]$. 
\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
\end{itemize}

@@

\subsection{Relation to \systemf}

To establish a correspondence between \fexists and standard unpolarized systems,
we present a translation between \fexists and \systemf. The translation of
\systemf to \fexists is given in \cref{fig:polarization}, and the backward translation
is given in \cref{fig:depolarization}.

\begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    [[|↓iN|]] &= [[|iN|]] \\
    [[|↑iP|]] &= [[|iP|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|α±|]] &= [[a]] \\
    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type depolarization}
  \label{fig:depolarization}
\end{figure}

\begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \qquad
  \begin{aligned}
    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
  \end{aligned}
  \end{equation*}
  \caption{Type polarization}
  \label{fig:polarization}
\end{figure}

### 2024-02-13 oopsla24/tex/03-declarative-system.tex
wip

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.
  The difference is that the former rule is for the 
  \emph{unannotated} let binder, and thus
  it requires the resulting type of application to be unique
  (up to equivalence), so that the type of the bound variable $[[x]]$
  is known before it is put into the context.
  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants throughout the algorithm to ensure
  the decidability of the produced unification task.
  In \cref{sec:proof-subtyping}, we discuss these invariants in
  more detail, and in \cref{sec:weakening-invariant} we suggest 
  a way to relax them.

>>>>

\paragraph{Applicative Let Binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

  The former rule infers the type of an \emph{unannotated} let binder, and thus
  it requires the resulting type of application to be \emph{the principal type},
  so that the type we assign to the bound variable $[[x]]$ is determined.
  In this context, principality means minimality. In other words, 
  $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
  any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
  is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.

  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants throughout the algorithm to ensure
  the decidability of the produced unification task.
  In \cref{sec:proof-subtyping}, we discuss these invariants in
  more detail, and in \cref{sec:weakening-invariant} we suggest 
  a way to relax them.

### 2024-02-13 oopsla24/tex/04-algorithm.tex
wip

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
    % Define nodes
    \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
    \node[right=1.2cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
    \node[below=0.8cm of 2] (7) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[below=0.3cm of 7] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
    \draw[->] ([yshift=+.2cm]1.east) to ([yshift=+.2cm]2.west);
    \draw[->, bend right=20] (3) to (1);
    \draw[->] (2) to (3);
    \draw[->] (2) to (4);
    \draw[->] (2) to (7);
    \draw[->] (3) to (4);
    \draw[->] (1.south west) |- (5);
    \draw[->] (3) |- (5);
    \draw[->] (2) to (6);
  \end{tikzpicture}  
  \caption{Dependency graph of the typing algorithm}
  \label{fig:alg-typing-graph}
\end{figure}

@@

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
  However, since there is no annotation, 
  instead of checking
  that the inferred type $[[↑uQ]]$ 
  can be a subtype of the annotation,
  the algorithm checks that $[[↑uQ]]$ is unique. 
  As we prove, uniqueness means that all the algorithmic variables of 
  $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$.
  This way, uniqueness is checked by the combination of
  $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
  Together, these two premises guarantee that the only 
  possible instantiation of $[[uQ]]$ is $[[ [uσ]uQ ]]$.

@@

\subsection{Constraint Singularity}
\label{sec:constraint-singularity}

The singularity algorithm is used to check that the inferred type of an
application is unique. In this case, the annotation in the applicative let
binder can be omitted. In the algorithmic system, the uniqueness means that the
\emph{constraint set} produced by the application inference has a single
substitution satisfying it.

>>>>

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
    % Define nodes
    \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
    \node[right=1.2cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
    \node[below=0.6cm of 2] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[below=0.5cm of 7] (8) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \node[below=0.9cm of 6] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
    \draw[->] ([yshift=+.2cm]1.east) to ([yshift=+.2cm]2.west);
    \draw[->, bend right=20] (3) to (1);
    \draw[->] (2) to (3);
    \draw[->] (2) to (4);
    \draw[->] (2) to (7);
    \draw[->] (3) to (4);
    \draw[->] (7) to (8);
    \draw[->] (1.south west) |- (5);
    \draw[->] (3) |- (5.south east);
    \draw[->] (2) to (6);
  \end{tikzpicture}  
  \caption{Dependency graph of the typing algorithm}
  \label{fig:alg-typing-graph}
\end{figure}

@@

  Rule \ruleref{\ottdruleATAppLetLabel}
  works similarly to \ruleref{\ottdruleATAppLetAnnLabel}.
  However, since no annotation is given,
  the algorithm must ensure that the inferred $[[uQ]]$
  has the `canonical' minimal instantiation.
  To find it, it makes a call to the minimal instantiation algorithm 
  (\cref{sec:minimal-instantiation})
  that finds the substitution that satisfies the inferred constraints $[[SC]]$ and
  instantiates $[[uQ]]$ to the minimal (among other such instantiations)
  type $[[ [uσ]uQ ]]$.

@@

\subsection{Minimal Instantiation and Constraint Singularity}
\label{sec:constraint-singularity}

Multiple types $[[iM]]$ can be inferred for a type application:
$[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ but only one \emph{principal} 
type should be chosen for a variable in an unannotated let binder.
Declaratively, we require the principal type $[[iP]]$ to be
minimal among other all types $[[iP']]$ that can be inferred
for the application $[[Γ ; Φ ⊢ iN ● args ⇒> ↑iP']]$.
Algorithmically, the inference returns an
\emph{algorithmic} type $[[uP]]$ 
together with a set of necessary and sufficient constraints $[[SC]]$
that any instantiation of $[[uP]]$ must satisfy.
This way, to guarantee principality, we must minimize the instantiation of $[[uP]]$,
while ensuring that it satisfies the given constraints $[[SC]]$.

The minimal instantiation judgment `$[[Γ ⊢ uP SC minby uσ]]$'
means that $[[uσ]]$ is the minimum among all the substitutions satisfying $[[SC]]$,
that is $[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
instantiation of $[[uP]]$. The algorithm is given in \cref{fig:minimal-instantiation}.
First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
Then it considers two cases. If the type is an algorithmic variable $[[α̂⁺]]$
restricted by a constraint $[[(α̂⁺ :≥ iP) ∊ SC]]$,
its minimal instantiation is the type $[[nf(iP)]]$, 
and \ruleref{\ottdruleSINGPUvarLabel} is applied.
Otherwise, \ruleref{\ottdruleSINGSingLabel} is applied,
meaning that all the instantiations of the type must be equivalent.
The algorithm ensures that by two premises:
\begin{enumerate*}
  \item[(i)] checking that all the algorithmic variables of the type are restricted by the constraint set; and
  \item[(ii)] building the substitution that satisfies the constraint set on these variables 
    (and simultaneously checking that such substitution is unique up-to-equivalence)
    using the constraint singularity algorithm.
\end{enumerate*}

\begin{figure}[h]
  \hfill\\
  \ottdefnMININSTLabeled
  \caption{Minimal Instantiation}
  \label{fig:minimal-instantiation}
\end{figure}

The singularity algorithm 
performs two tasks: it checks that a
\emph{constraint set} has a single substitution satisfying it, 
and if it does, it builds this substitution.

### 2024-02-13 oopsla24/tex/05-proof.tex
wip

\subsection{Singularity}
\label{sec:singularity-proof}

The rule \ruleref{\ottdruleDTAppLetLabel} of the declarative inference
says that one can omit the type annotation of a let-binding if the inferred 
type of the bound application is unique. Accordingly, the algorithm,
that returns the constraint set, must be able to check that
the set of constraints has a single solution, and also to find this solution. 
This subroutine is called singularity, and its correctness is formulated as follows.

The soundness states that $[[SC singular with uσ]]$
implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
on the domain.  The completeness states that if all the 
$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.

>>>>

\subsection{Minimal Instantiation and Singularity}
\label{sec:singularity-proof}

\paragraph{Minimal Instantiation}
Algorithmic typing relies on the \emph{minimal instantiation} procedure.
For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
This way, the correctness of minimal instantiation is formulated as the following two lemmas.

\begin{lemma*}[Soundness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
    Then $[[Γ ⊢ uP SC minby uσ ]]$ implies that 
    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
    (\ie $[[Θ ⊢ uσ : SC]]$)
    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$ holds
\end{lemma*}

\begin{lemma*}[Completeness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
    satisfying $[[SC]]$, we have
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
    succeeds, specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
\end{lemma*}

The soundness is rather straightforwardly proved by induction on the 
judgment inference tree, relying on the soundness of the singularity procedure.
The proof of completeness is done by induction on the structure of the type $[[uP]]$.
It uses the fact that $[[ Γ ⊢ ∃nas.[uσ']uP ≥ ∃nas.[uσ]uP]]$ implies
$[[ Γ, nas ⊢ [uσ']uP ≥ [uσ]uP]]$.

\paragraph{Singularity}
The soundness of singularity states that $[[SC singular with uσ]]$
implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
on the domain.  The completeness states that if all the 
$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.
